Fine-scale Twitter Big Data Reveals Neighborhood Inequalities in Mental Health Effects of High Temperatures

### 2021-05-01 doc/temp-sentiment.tex

\documentclass{article}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage[a4paper, margin=1.25in]{geometry}
\usepackage{breakcites}
\usepackage{subcaption}
\usepackage{float}
\usepackage{textcomp}
\usepackage{amsmath}
\usepackage{textgreek}
\usepackage{authblk}
\usepackage{rotating}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{pdflscape}
\usepackage{lineno}
\usepackage{xcolor}
\usepackage[
  style=numeric,
  citestyle=numeric-comp,
  backend=biber,
  doi=true,
  natbib=true,
  sorting=none
]{biblatex}

@@

@@

\subsection{Overall Effect}
We examined the relationship between the sentiment expressed in tweets and the prevailing Wet Bulb Globe Temperature (WBGT), an indicator of heat stress that accounts for temperature, humidity, wind speed, and solar radiation.  Controlling for a variety of fixed effects, we found that, across all tweets, higher temperatures are associated with lower sentiment (See Fig. \ref{fig:wbgt}).  Sentiment is highest at a WBGT of 5 degrees Celsius (a dry bulb temperature typically around 12C/54F).  The greatest declines in sentiment are observed between 20-25 degrees WBGT (dry bulb 29-36C/84-97F).  Sentiment also declines with colder temperatures, but only slightly.
We ran models using multiple sentiment indicators, and found similar results no matter which sentiment metric was used (See Appendix).

@@

We then explored how the racial and income characteristics of neighborhoods moderate the relationship between WBGT and sentiment (See Fig \ref{fig:sub1}).  We found that neighborhood income strongly moderates the relationship between temperature and sentiment.  As temperatures increase to moderate heat of 20 WBGT (29C/84F), sentiment decreases for most neighborhoods, but increases in neighborhoods in the 95th income percentile.  These rich do not see decreases in sentiment until temperatures are above 20 WBGT (29C/84F), at which point sentiment decreases evenly across income ranges, still leaving a large gap in expressed sentiment from low to high-income neighborhoods.

We examined patterns of temperature and sentiment for neighborhoods that had a majority population of four broad racial categories - non-hispanic black, non-hispanic white, hispanic of any race, and other, which includes native american, multi-racial, and asian-american and pacific islander.  While people in neighborhoods of all racial compositions were affected by heat, people in majority black neighborhoods were much more affected than others  (See Fig \ref{fig:sub2}).  Relative to an optimum temperature of 5 WBGT (12C/54F), as temperatures increase to 30 WBGT (42C/108F), the sentiment of tweets in majority black neighborhoods decreases four-times as much as people in other neighborhoods.  Additionally, at levels of moderate heat from 10 WBGT (19C/66F) to 25 WBGT (36C/97F), people in majority hispanic neighborhoods have slightly lower sentiment than people in majority white or other neighborhoods, although this gap closes with high temperatures.

@@

Expressed sentiment is a widely used metric to assess mental health and well-being on social media.  While, a variety of related metrics can be used to quantify sentiment, we chose to use the VADER metric that was specifically designed for microblogs like twitter \cite{hutto2014vader}.  To give context to this metric, we compare the impacts of heat waves, defined as a change from 5 WBGT (12C/54F) to 25 WBGT (36C/97F), to the change in average sentiment from Saturday to Monday as well as the impacts on of the two most expensive hurricanes in the last decade in the United States (See Fig. \ref{fig:compare}.  Saturday and Monday are the highest and lowest days for both twitter sentiment and suicides, and over the course of our study period, Mondays had 23\% more suicides than Saturdays.  For the hurricanes, we compared the mean sentiment of counties affected by Hurricanes Harvey and Sandy on the day of landfall to the mean sentiment a week before.  These hurricanes are associated with mental health effects including Post-Traumatic Stress Disorder (PTSD) \cite{Schwartz2017Aug, Schwartz2018May}, and similar disasters have been associated with subsequent increases in suicides \cite{Krug1998Feb}.

@@

To further explore our findings, we examined inequalities in the impacts of higher temperatures on sentiment by Combined Statistical Areas (CSAs) and Metropolitan Statistical Areas (MSAs) with over 1 million people (hereafter: cities).  We examined the change in size of the gap in sentiment scores between both poor and rich neighborhoods, as well as black and non-black neighborhoods, as temperatures change from optimum temperatures of 5 WBGT (12C/54F) to 30 WBGT (42C/108F).

@@

\begin{figure}[H]
\centering
\begin{subfigure}{0.75\textwidth}
  \includegraphics[width=\linewidth]{../res/map_wbgt_income.png}
  \caption{}
  \label{fig:map1}
\end{subfigure}
\begin{subfigure}{0.75\textwidth}
  \includegraphics[width=\linewidth]{../res/map_wbgt_black.png}
  \caption{}
  \label{fig:map2}
\end{subfigure}
\caption{Inequalities in impacts of temperature on sentiment by CSAs and MSAs with over 1 million people.  The value shown is the predicted change in the size of the gap in vader score as temperatures increase from 5 WBGT to 30 WBGT between the 5th and 95 percentile for income (a) and the 95th and 5th percentile for percent black (b).  CSAs with a statistically significant effect ($p < 0.05$) are outlined in black.  For racial inequalities (b), only MSAs and CSAs where black people make up >5\% of the population are included.}
\label{fig:test}
\end{figure}

@@

\begin{figure}[H]
  \centering
  \includegraphics[width=0.75\linewidth]{../res/ts.png}
  \caption{Effects of rising temperatures on sentiment by hour of the day, with a 95\% confidence interval.  The value shown is the predicted change in vader score for a 25-degree increase in temperatures.}
  \label{fig:ts-wbgt}
\end{figure}

@@

However, one limitation of data from twitter is that a tweet is not a precise measurement of a mental health outcome, but rather only a rough indicator a user's mental state based on the vocabulary in the tweet.  Moreover, while sentiment analysis algorithms have gotten increasingly sophisticated at estimating the mood in a body of text, accounting for negations (``\textit{not} good"), punctuation (``Good!!!"), emoticons and emojis (":-) or d grin

% This could honestly be its own appendix section
  The change from optimum temperature of 5WBGT (12C/54F) to a heatwave of 25WBGT (36C/97F) is associated with an overall decrease in the vader sentiment score of 0.01.  This is similar to the degree of change in sentiment over the course of a week from a Monday nadir (0.1268) to a Saturday peak (0.1372).  Additionally, over the study period from 2009-2019, suicides followed a very similar weekly pattern, peaking on Mondays and declining over the course of the week until Saturday.  From Saturday to Monday, suicides increase 23\%.  Thus, while research on linkages between sentiment and other mental health outcomes like suicides and hospitalizations is still nascent, there are clear similarities in patters of sentiment at some time scales.  Moreover, both suicides and sentiment are affected by higher temperatures.  Thus, while twitter sentiments are the only mental health indicator available at fine enough spatial and temporal scales to examine neighborhood and hourly effects, these changes in sentiment are representative of real human suffering, and are also very likely indicative of medical outcomes like suicides and hospitalizations.

Recent research has suggested that the impacts of heat on sleep quality may play a large role in the observed mental health effects of heat, 
\section{Data}
A proposed list of tables and figures:
Figures for location of USA Tweets (to show aggregation and coverage).
Figures of income blocks and temperature extremes.
Table 1. Variable descriptions for what's included in our models.
Table 2. Descriptive statistics for datasets/variables we include.
Table 3. Model results with variables included in the models coefficients, SEs, p-values, sample size.
Appendix table. Sentiment scores across all types we considered.

@@

[NEED SOME TEXT ABOUT GOOGLE EARTH ENGINE AND MICROSOFT AI FOR EARTH] We conducted all data analyses on Microsoft AI for Earth's online cloud-computing environment using a combination of existing data analysis packages and scripts we developed. We used GitHub as a repository for all scripts we used in this project (link to the Git repo). To run ABC analysis, we used Microsoft's CBA package. To run XYZ analysis, we used Microsoft's ZYX package.

%%KEY POINTS TO DISCUSS

% Link between twitter mood and other very bad outcomes (crime, suicide)

% Baselines: this could also be interpreted as: poor are less affected by the cold\textbf{:}

% Caveats:
    % Uncertainties from census blocks (low income, young areas).  This is why we used a linaer effect instead of a binned effect.
    % We cant speak to types of mental heath that are worsening - depression, anxiety, and post-traumatic stress, etc.  Just a general overview

% More nuanced interpretation: optimal temperature for poor & minority neighborhoods are much lower.

\section{Conclusion}

>>>>

\documentclass{article}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage[a4paper, margin=1.25in]{geometry}
\usepackage{breakcites}
\usepackage{subcaption}
\usepackage{float}
\usepackage{textcomp}
\usepackage{amsmath}
\usepackage{textgreek}
\usepackage{authblk}
\usepackage{rotating}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{pdflscape}
\usepackage{lineno}
\usepackage{xcolor}
\usepackage[
  style=numeric,
  citestyle=numeric-comp,
  backend=biber,
  doi=true,
  natbib=true,
  sorting=none
]{biblatex}
\usepackage{scalerel,xparse}

@@

% To be able to use emojis
\NewDocumentCommand\emojismile{}{
    \scalerel*{
        \includegraphics{smiling-face-with-smiling-eyes_1f60a.png}
    }{X}
}

@@

\subsection{Overall Effect}
We examined the relationship between the sentiment expressed in tweets and the prevailing Wet Bulb Globe Temperature (WBGT), an indicator of heat stress that accounts for temperature, humidity, wind speed, and solar radiation.  Controlling for a variety of fixed effects, we found that, across all tweets, higher temperatures are associated with lower sentiment (See Fig. \ref{fig:wbgt}).  Sentiment is highest at a WBGT of 5\textdegree Celsius (a dry bulb temperature typically around 12\textdegree C/54\textdegree F).  The greatest declines in sentiment are observed between 20\textdegree -25\textdegree C WBGT (dry bulb 29\textdegree-36\textdegree C/84\textdegree-97\textdegree F).  Sentiment also declines with colder temperatures, but only slightly.  We ran models using multiple sentiment indicators, and found similar results no matter which sentiment metric was used (See Appendix).

@@

We then explored how the racial and income characteristics of neighborhoods moderate the relationship between WBGT and sentiment (See Fig \ref{fig:sub1}).  We found that neighborhood income strongly moderates the relationship between temperature and sentiment.  As temperatures increase to moderate heat of 20\textdegree C WBGT (29\textdegree C/84\textdegree F), sentiment decreases for most neighborhoods, but increases in neighborhoods in the 95th income percentile.  These rich do not see decreases in sentiment until temperatures are above 20\textdegree C WBGT (29\textdegree C/84\textdegree F), at which point sentiment decreases evenly across income ranges, still leaving a large gap in expressed sentiment from low to high-income neighborhoods.

We examined patterns of temperature and sentiment for neighborhoods that had a majority population of four broad racial categories - non-hispanic black, non-hispanic white, hispanic of any race, and other, which includes native american, multi-racial, and asian-american and pacific islander.  While people in neighborhoods of all racial compositions were affected by heat, people in majority black neighborhoods were much more affected than others  (See Fig \ref{fig:sub2}).  Relative to an optimum temperature of 5\textdegree C WBGT (12\textdegree C/54\textdegree F), as temperatures increase to 30\textdegree C WBGT (42\textdegree C/108\textdegree F), the sentiment of tweets in majority black neighborhoods decreases four-times as much as people in other neighborhoods.  Additionally, at levels of moderate heat from 10\textdegree C WBGT (19\textdegree C/66\textdegree F) to 25\textdegree C WBGT (36\textdegree C/97\textdegree F), people in majority hispanic neighborhoods have slightly lower sentiment than people in majority white or other neighborhoods, although this gap closes with high temperatures.

@@

Expressed sentiment is a widely used metric to assess mental health and well-being on social media.  While, a variety of related metrics can be used to quantify sentiment, we chose to use the VADER metric that was specifically designed for microblogs like twitter \cite{hutto2014vader}.  To give context to this metric, we compare the impacts of heat waves, defined as a change from 5\textdegree C WBGT (12\textdegree C/54\textdegree F) to 25\textdegree C WBGT (36\textdegree C/97\textdegree F), to the change in average sentiment from Saturday to Monday as well as the impacts on of the two most expensive hurricanes in the last decade in the United States (See Fig. \ref{fig:compare}.  Saturday and Monday are the highest and lowest days for both twitter sentiment and suicides, and over the course of our study period, Mondays had 23\% more suicides than Saturdays.  For the hurricanes, we compared the mean sentiment of counties affected by Hurricanes Harvey and Sandy on the day of landfall to the mean sentiment a week before.  These hurricanes are associated with mental health effects including Post-Traumatic Stress Disorder (PTSD) \cite{Schwartz2017Aug, Schwartz2018May}, and similar disasters have been associated with subsequent increases in suicides \cite{Krug1998Feb}.

@@

To further explore our findings, we examined inequalities in the impacts of higher temperatures on sentiment by Combined Statistical Areas (CSAs) and Metropolitan Statistical Areas (MSAs) with over 1 million people (hereafter: cities).  We examined the change in size of the gap in sentiment scores between both poor and rich neighborhoods, as well as black and non-black neighborhoods, as temperatures change from optimum temperatures of 5\textdegree C WBGT (12\textdegree C/54\textdegree F) to 30\textdegree C WBGT (42\textdegree C/108\textdegree F).

@@

\begin{figure}[H]
\centering
\begin{subfigure}{0.75\textwidth}
  \includegraphics[width=\linewidth]{../res/map_wbgt_income.png}
  \caption{}
  \label{fig:map1}
\end{subfigure}
\begin{subfigure}{0.75\textwidth}
  \includegraphics[width=\linewidth]{../res/map_wbgt_black.png}
  \caption{}
  \label{fig:map2}
\end{subfigure}
\caption{Inequalities in impacts of temperature on sentiment by CSAs and MSAs with over 1 million people.  The value shown is the predicted change in the size of the gap in vader score as temperatures increase from 5WBGT to 30WBGT between the 5th and 95 percentile for income (a) and the 95th and 5th percentile for percent black (b).  CSAs with a statistically significant effect ($p < 0.05$) are outlined in black.  For racial inequalities (b), only MSAs and CSAs where black people make up >5\% of the population are included.}
\label{fig:test}
\end{figure}

@@

\begin{figure}[H]
  \centering
  \includegraphics[width=0.5\linewidth]{../res/ts.png}
  \caption{Effects of rising temperatures on sentiment by hour of the day, with a 95\% confidence interval.  The value shown is the predicted change in vader score for a 25-degree increase in temperatures.}
  \label{fig:ts-wbgt}
\end{figure}

@@

However, one limitation of data from twitter is that a tweet is not a precise measurement of a discrete mental health outcome and only provides a rough indicator a user's mental state based on the vocabulary in the tweet.  Moreover, while sentiment analysis algorithms have gotten increasingly sophisticated at estimating the mood in a body of text, accounting for negations (``\textit{not} good"), punctuation (``Good!!!"), emoticons and emojis (``:-)" or ``\emojismile"), and degree modifiers (``very excellent" or ``kind of crappy") \cite{hutto2014vader}, the observed sentiment in a tweet is highly variable and is mostly affected by factors like current events and the user's personal life, with local weather conditions having only a small impact.  We overcome this limitation by using an extremely large dataset of nearly 250 million tweets, allowing us to measure population-level changes in sentiment.

We found that the change from optimum temperature of 5\textdegree C WBGT (12\textdegree C/54\textdegree F) to a heatwave of 25\textdegree C WBGT (36\textdegree C/97\textdegree F) is associated with an overall decrease in the vader sentiment score of 0.01.  This is similar to the degree of change in sentiment over the course of a week from a Monday nadir (0.1268) to a Saturday peak (0.1372), a change in sentiment associated with a suicides increase of 23\% over the course of the study period \cite{CDC2021}.  Thus, while research on linkages between sentiment and other mental health outcomes like suicides and hospitalizations is still nascent, there are clear similarities in patters of sentiment and suicides at some time scales.  Moreover, both suicides and sentiment are affected by higher temperatures \cite{baylis_weather_2018, Burke2018Aug}.  Thus, while twitter sentiment is the only mental health indicator available at fine enough spatial and temporal scales to examine neighborhood and hourly effects, these changes in sentiment are representative of real human suffering, and are also very likely indicative of medical outcomes like suicides and hospitalizations, although this analysis cannot speak to those outcomes specifically.

We examined our results relative to an optimal baseline temperature of 5\textdegree C WBGT (12\textdegree C/54\textdegree F), as this was the temperature at which the highest sentiment scores were observed in the aggregated analysis (See Fig. \ref{fig:wbgt}).  However, after examining heterogeneities by neighborhood race and income, we find different optima for different racial groups and income levels.  For example, the optimum temperature for poor neighborhoods is 0\textdegree C WBGT (6\textdegree C/43\textdegree F) while the optimum temperature for rich neighborhoods is 20\textdegree C WBGT (29\textdegree C/84\textdegree F).  These different varying optima show that even mild heat can affect human well-being, and the only reason that sentiment peaks as high as 20\textdegree C WBGT (29\textdegree C/84\textdegree F) in rich neighborhoods is because the rich are more likely to have air conditioning in their homes, transportation, and work spaces, as well as much greater choice in when they are outside and in the activities they perform outside.

We found that majority black neighborhoods are much more affected by heat than neighborhoods with a majority of any other race.  This is surprising, given that Hispanics can be marginalized in both income and housing.  This finding may be because the racial category of Hispanic is more broad and encompasses many more groups with more diverse histories and income levels than black Americans.  Additionally, there are many other marginalized groups in America, particularly indigenous people, that we did not have sufficient data to examine specifically with respect to heat and mental health.  Finally, in modeling the combined effects of race and income in moderating the impacts of heat on mental health, we find separate effect for these two variables (See Appendix).  Neither race or income alone account for all the heterogeneity in vulnerability, and neighborhoods that are poor and black are more affected by than than neighborhoods that are just poor or just black.

In our spatial analysis, we found that most cities exhibited both racial and income inequalities in the vulnerability of mental health to heat, with noticeable regional patterns in the strength of the effect.  Generally, the mid-west and mid-atlantic had the strongest inequalities, and the south-west had the weakest inequalities and an occasional inverted effect.  Because we used Wet Bulb Globe Temperature, our metric accounts for how differences in humidity and wind speed affect the human body's ability to cool itself through perspiration.  Thus, this regional differences in vulnerability are likely more due to cultural or infrastructural differences than to different environmental conditions.

Recent research has suggested that the impacts of heat on sleep quality may play a large role in the observed mental health effects of heat \cite{Obradovich2017May, Mullins2019Dec}, and our temporal analysis was able to examine this hypothesis.  While our estimates were more uncertain at night due to a lower volume of tweets, we found much stronger effects of heat on expressed sentiment in the early morning, adding weight to the sleep quality pathway linking heat and mental health that other authors have found.  This suggests that efforts to improve mental health during heat waves may have the largest impact by focusing on providing electricity and air conditioning at night.

While these findings are robust to different model specifications and sentiment metrics, there are some caveats.  For one, we were only able to locate the tweets within the census block where the tweet was sent - we did not to infer where the person sending the tweets typically lived or where they had been.  Many low-income and black people commute during the day to work in the service sector in higher-income areas, so these results may in fact under-estimate the impacts of higher temperatures on mental health.  Moreover, census block income levels can give inaccurate estimates of the well-being of the people who are typically found in census blocks.  For example, some public spaces are estimated to have very low income levels even though people from a variety of income levels may inhabit those spaces throughout the day.  Additionally, neighborhoods of young college students are also typically estimated to have low income levels, even through college students are wealthier than the average American.  Again, these issues mean we are likely under-estimating the true effect of heat on mental health.

While climate change will have widespread and severe impacts on human well-being, these impacts are will be highly unevenly distributed.  People with more money, access to aid and infrastructure, and who belong to ethnic groups in power are less affected by climate shocks and natural disasters \cite{bullard2012wrong}.  While the physical health impacts of these shocks are more visible and easy to measure, there is no reason to believe that the mental health impacts of climate change will not also be highly unevenly distributed.  Thus, there are strong theoretical priors behind the hypothesis that low-income and marginalized people are more vulnerable to the mental health impacts of higher temperatures, even though previous work at coarse spatial scales has not found such an effect.  By using fine-scale twitter data, we showed that there are indeed stark difference in the mental health impacts of heat among neighborhoods in America.

\section{Data}
A proposed list of tables and figures:
Figures for location of USA Tweets (to show aggregation and coverage).
Figures of income blocks and temperature extremes.
Table 1. Variable descriptions for what's included in our models.
Table 2. Descriptive statistics for datasets/variables we include.
Table 3. Model results with variables included in the models coefficients, SEs, p-values, sample size.
Appendix table. Sentiment scores across all types we considered.

@@

[NEED SOME TEXT ABOUT GOOGLE EARTH ENGINE AND MICROSOFT AI FOR EARTH] We conducted all data analyses on Microsoft AI for Earth's online cloud-computing environment using a combination of existing data analysis packages and scripts we developed. We used GitHub as a repository for all scripts we used in this project (link to the Git repo). To run ABC analysis, we used Microsoft's CBA package. To run XYZ analysis, we used Microsoft's ZYX package. 
\section{Conclusion}
